---
title: "How To Use CGMissingDataR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How To Use CGMissingDataR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Overview

CGMissingDataR imputes glucose values that are already missing in continuous
 glucose monitoring (CGM) data. The main user-facing function is:

```r
run_missing_glucose_imputation()
```

The function is designed for real missing glucose values. It returns a single
 data frame with the original glucose column left unchanged and a new completed
 glucose column named `imputed_glucose_value`.

The core workflow is:

1. read a data frame or CSV file;
2. create or reuse a `TimeSeries` column;
3. encode `SEX` when present;
4. create internal lag and rolling-mean glucose features;
5. impute the target and feature matrix;
6. choose `MICE+ARIMA` or `MICE+XGBoost` from the observed missing rate;
7. remove internal-only engineered features before returning the data frame.

# Installation

Install the CRAN release with:

```r
install.packages("CGMissingDataR")
```

Install the development version with:

```r
install.packages("devtools")
devtools::install_github("ZhangLabUKY/CGMissingDataR")
```

Load the package:

```{r setup}
library(CGMissingDataR)
```

# Example data

`CGMExampleData` is a small multi-subject CGM data set included with the
 package. It contains a subject identifier, raw timestamp column, glucose column,
 age, and HbA1c.

```{r example-data}
data("CGMExampleData")

summary_table <- data.frame(
  Rows = nrow(CGMExampleData),
  Columns = ncol(CGMExampleData),
  Subjects = length(unique(CGMExampleData$USUBJID)),
  MissingGlucose = sum(is.na(CGMExampleData$LBORRES)),
  MissingPercent = round(mean(is.na(CGMExampleData$LBORRES)) * 100, 1)
)

summary_table
head(CGMExampleData)
```

The example data intentionally does not include `TimeSeries`. The imputation
 function creates that column internally from the raw `Time` column.

# Required input columns

At minimum, the imputation function needs:

| Role | Argument | Example column |
|---|---|---|
| Glucose value to impute | `target_col` | `LBORRES` |
| Subject identifier | `id_col` | `USUBJID` |
| Raw timestamp | `time_col` | `Time` |
| Additional predictors | `feature_cols` | `AGE`, `hba1c` |

The target column may contain missing values. Predictor columns should be numeric
 or coercible to numeric. The `SEX` column, when present, is internally encoded
 as `M = 1` and `F = 0`.

# Basic real-imputation workflow

For the CRAN-safe R-native path, use `imputer_backend = "mice"`.

```{r basic-imputation}
impute_out <- suppressWarnings(
  run_missing_glucose_imputation(
    CGMExampleData,
    target_col = "LBORRES",
    feature_cols = c("AGE", "hba1c"),
    id_col = "USUBJID",
    time_col = "Time",
    imputer_backend = "mice",
    prefer_cgmanalyzer_equal_interval = FALSE,
    xgb_nrounds = 5
  )
)
```

The result is a data frame:

```{r output-shape}
class(impute_out)
nrow(impute_out)
names(impute_out)
```

The most important returned columns are:

| Column | Meaning |
|---|---|
| Original target column, e.g. `LBORRES` | The original glucose column. Values originally missing remain `NA`. |
| `TimeSeries` | Numeric elapsed time feature derived from the timestamp column. |
| `imputed_glucose_value` | Completed glucose values after imputation. |
| `imputation_method` | Final method used: `MICE+ARIMA` or `MICE+XGBoost`. |
| `missing_rate` | Original missing rate of the target column. |

```{r output-preview}
head(impute_out[c(
  "USUBJID",
  "Time",
  "TimeSeries",
  "LBORRES",
  "imputed_glucose_value",
  "imputation_method",
  "missing_rate"
)])
```

The original target column is not overwritten:

```{r original-target-unchanged}
sum(is.na(CGMExampleData$LBORRES))
sum(is.na(impute_out$LBORRES))
sum(is.na(impute_out$imputed_glucose_value))
```

Inspect only the rows where glucose was originally missing:

```{r missing-row-preview}
missing_rows <- is.na(impute_out$LBORRES)
head(impute_out[missing_rows, c(
  "USUBJID",
  "Time",
  "LBORRES",
  "imputed_glucose_value",
  "imputation_method"
)])
```

# How the method is selected

The function automatically chooses the final imputation model from the target
 missing rate:

- if missing rate is less than or equal to `use_arima_if_missing_leq`, the final
  method is `MICE+ARIMA`;
- otherwise, the final method is `MICE+XGBoost`.

The default threshold is `0.05`.

```{r method-selection}
unique(impute_out$missing_rate)
unique(impute_out$imputation_method)
```

For data sets with more than 5% target missingness, the default final method is
 usually `MICE+XGBoost`. For lower missingness, the function uses segmentwise
 ARIMA when enough subject-level history is available.

# Time handling

The function accepts common timestamp formats, including colon-separated,
 hyphen-separated, slash-separated, ISO-style, and `POSIXct` inputs.

Examples of accepted character formats include:

```r
"2020:01:16:00:00"
"2020-01-16 00:00:00"
"2020/01/16 00:00:00"
"01/16/2020 00:00"
"2020-01-16T00:00:00"
```

If a non-empty `TimeSeries` column is already present, the function reuses it.
 Otherwise, it creates `TimeSeries` from the timestamp column.

For quieter reproducible examples, this vignette uses:

```r
prefer_cgmanalyzer_equal_interval = FALSE
```

That bypasses `CGManalyzer::equalInterval.fn()` and uses elapsed minutes by
 subject. Users who want the CGManalyzer equal-interval behavior can leave the
 default value as `TRUE`.

# Internal engineered features

The workflow creates lag and rolling-mean features before imputation. These
 features are used internally by the imputer and final model, but they are not
 included in the returned data frame.

```{r internal-feature-check}
grep("^lag[0-9]+$|^rollmean$", names(impute_out), value = TRUE)
```

This should return an empty character vector because those features are internal
 implementation details.

# Optional Python-compatible backend

For closest agreement with the Python reference workflow, use:

```r
imputer_backend = "sklearn"
```

In that mode, the function sends the input data frame to Python through
 `reticulate`. Python then performs preprocessing and imputation with:

- `pandas` for data-frame operations;
- `scikit-learn` for `IterativeImputer`;
- `statsmodels` for ARIMA;
- Python `xgboost` for XGBoost regression.

The completed pandas data frame is then converted back to R.

## Installing optional Python dependencies

Install `reticulate` in R:

```r
install.packages("reticulate")
```

Declare the Python dependencies before running the Python backend:

```r
reticulate::py_require(c(
  "numpy",
  "pandas",
  "scikit-learn",
  "statsmodels",
  "xgboost"
))
```

Then call the function with `imputer_backend = "sklearn"`:

```{r python-backend-example, eval = FALSE}
out_py <- run_missing_glucose_imputation(
  CGMExampleData,
  target_col = "LBORRES",
  feature_cols = c("AGE", "hba1c"),
  id_col = "USUBJID",
  time_col = "Time",
  imputer_backend = "sklearn",
  xgb_nrounds = 5
)

head(out_py[c(
  "USUBJID",
  "Time",
  "LBORRES",
  "imputed_glucose_value",
  "imputation_method"
)])
```

The Python backend is optional. It is not required for package installation or
 for building this vignette.

# Choosing a backend

| Backend | Use case | Notes |
|---|---|---|
| `mice` | Default R-native workflow | CRAN-safe and does not require Python. |
| `sklearn` | Closest Python-compatible workflow | Requires `reticulate` and Python packages. |

Use `mice` for simple installation and CRAN-safe examples. Use `sklearn` when
 comparing with the Python reference workflow or when you want Python libraries
 to perform the full strict path.

# Exporting results

Set `export = TRUE` to write the returned imputed data frame to a timestamped
 CSV file in the current working directory.

```{r export-example, eval = FALSE}
out <- run_missing_glucose_imputation(
  CGMExampleData,
  target_col = "LBORRES",
  feature_cols = c("AGE", "hba1c"),
  id_col = "USUBJID",
  time_col = "Time",
  imputer_backend = "mice",
  export = TRUE
)
```

The function still returns the data frame invisibly to the assignment target.

# Troubleshooting

## Timestamp parsing errors

If you see an error such as:

```r
Some timestamp values could not be parsed
```

check the values in your timestamp column:

```r
head(unique(your_data$Time))
```

Use a standard format such as `YYYY-mm-dd HH:MM:SS`, `YYYY:mm:dd:HH:MM`, or a
 `POSIXct` column.

## Python module errors

If the Python backend reports a missing module such as `sklearn`, remember that
 the package is installed as `scikit-learn` but imported as `sklearn`.

```r
reticulate::py_require(c("scikit-learn", "pandas", "statsmodels", "xgboost"))
```

If Python was already initialized before declaring requirements, restart R and
 run the call again.

## Warnings from `mice`

Small or highly collinear data sets can cause `mice` to report logged events.
 This is common with tiny examples and does not necessarily indicate failure.
 With real data, inspect those warnings to decide whether columns should be
 removed, recoded, or simplified.

# Session information

```{r session-info}
utils::sessionInfo()
```
