---
title: "Using the CGMissingDataR Shiny App"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the CGMissingDataR Shiny App}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Overview

CGMissingDataR includes an optional Shiny app for interactive missing glucose
imputation. The app is a point-and-click interface around the main package
function:

```r
run_missing_glucose_imputation()
```

The app is useful when users want to:

- upload a CSV file without writing R code;
- choose the target glucose, subject ID, timestamp, and feature columns from a
  user interface;
- load built-in example data sets for demonstration;
- inspect the observed missingness before running imputation;
- run the imputation workflow;
- preview only the rows where glucose was originally missing and then imputed;
- download the completed data as a CSV file.

The Shiny app does not implement a separate imputation algorithm. It calls
`run_missing_glucose_imputation()` internally and returns the same type of
completed data frame as the command-line workflow.

# Installation

Install CGMissingDataR from CRAN with:

```r
install.packages("CGMissingDataR")
```

The app requires the optional R package `shiny`. If Shiny is not already
installed, install it with:

```r
install.packages("shiny")
```

Then load the package:

```{r setup}
library(CGMissingDataR)
```

# Launching the app

Launch the app with:

```{r launch-app, eval = FALSE}
run_cgmissingdata_app()
```

During package development, after running `devtools::load_all()`, the same
launcher can be used:

```{r launch-app-development, eval = FALSE}
devtools::load_all()
run_cgmissingdata_app()
```

The app is bundled inside the installed package, typically under:

```r
system.file(
  "shiny",
  "cgm_imputation_app",
  package = "CGMissingDataR"
)
```

Users normally do not need to access this directory directly. The
`run_cgmissingdata_app()` launcher finds it automatically.

# Input options

The app provides two ways to load data.

## Upload a CSV file

Use the **Browse** button to upload a CSV file containing CGM data. The file
should contain, at minimum, columns corresponding to:

| Role | Example column | App selector |
|---|---:|---|
| Subject identifier | `USUBJID` | Subject ID column |
| Glucose value | `LBORRES` | Target glucose column |
| Timestamp | `Time` | Timestamp column |
| Additional predictors | `AGE`, `hba1c` | Feature columns |

After the file is uploaded, the app displays a preview of the uploaded data and
populates the column-selection controls.

## Load built-in example data

The app can also load built-in example data sets for demonstration. These are
useful for quickly showing how the workflow behaves without requiring users to
upload their own data.

The example data sets are intended to include:

| Example data | Description |
|---|---|
| `CGMExmplDat5Pct` | Example CGM data with about 5% missing glucose values. |
| `CGMExmplDat10Pct` | Example CGM data with about 10% missing glucose values. |


After selecting an example data set and clicking **Load example data**, the app
uses that data set exactly as if it had been uploaded by the user.

# Selecting columns

Once data are loaded, select the columns that map to the imputation function.

## Target glucose column

Choose the glucose column with missing values to impute. In the included example
data, this is usually:

```r
LBORRES
```

The original target column is preserved in the returned data. Values that were
originally missing remain `NA` in this original column. Completed glucose values
are written to a new column named:

```r
imputed_glucose_value
```

## Subject ID column

Choose the column identifying each subject or participant. In the example data,
this is usually:

```r
USUBJID
```

The subject ID is used for sorting, lag feature creation, rolling-mean feature
creation, and subject-level time handling.

## Timestamp column

Choose the raw timestamp column. In the example data, this is usually:

```r
Time
```

The imputation function creates or reuses a numeric `TimeSeries` column from the
timestamp values. Common timestamp formats are supported, including
colon-separated, hyphen-separated, slash-separated, ISO-style, and `POSIXct`
values.

## Feature columns

Choose additional predictor columns. In the example data, these commonly include:

```r
AGE
hba1c
```

Feature columns should be numeric or coercible to numeric. If a `SEX` column is
present, the underlying function can encode it internally.

# Missingness summary card

The app includes a missingness summary card beside the uploaded data preview.
After a target glucose column is selected, this card shows:

- the percentage of missing values in the target column;
- the number of missing rows;
- the total number of rows;
- a warning style when missingness is greater than the chosen threshold, such as
  20%.

This card is intended as a quick data-quality check before running the
imputation workflow. Higher missingness does not necessarily mean imputation
cannot be run, but users should interpret results carefully when a large portion
of the target glucose column is missing.

# Backend selection

The app supports the same backends as `run_missing_glucose_imputation()`.

| Backend | Description | Recommended use |
|---|---|---|
| `mice` | R-native backend using the R package `mice`. | Default, CRAN-safe workflow. |
| `sklearn` | Optional Python-compatible backend through `reticulate`. | Closest agreement with the Python reference workflow. |

## MICE backend

The default backend is:

```r
imputer_backend = "mice"
```

This backend does not require Python and is the safest choice for most users.
It is also the backend used in CRAN-safe examples and tests.

## Optional sklearn backend

The optional Python-compatible backend is:

```r
imputer_backend = "sklearn"
```

This path sends the data frame to Python through `reticulate`. Python then uses:

- `pandas` for data-frame operations;
- `scikit-learn` for `IterativeImputer`;
- `statsmodels` for ARIMA;
- Python `xgboost` for XGBoost regression.

To use the Python backend, install `reticulate` and declare the Python
requirements before launching or running the app:

```{r python-requirements, eval = FALSE}
install.packages("reticulate")

reticulate::py_require(c(
  "numpy",
  "pandas",
  "scikit-learn",
  "statsmodels",
  "xgboost"
))
```

The Python backend is optional. It is not required for installing or loading the
package.

# Running imputation

After loading data and selecting columns, click **Run imputation**.

Internally, the app calls code equivalent to:

```{r app-equivalent-call, eval = FALSE}
out <- run_missing_glucose_imputation(
  data = uploaded_data,
  target_col = selected_target_col,
  feature_cols = selected_feature_cols,
  id_col = selected_id_col,
  time_col = selected_time_col,
  imputer_backend = selected_backend,
  use_arima_if_missing_leq = selected_threshold,
  xgb_nrounds = selected_xgb_rounds,
  seed = selected_seed,
  export = FALSE
)
```

The returned object is a data frame. The most important columns are:

| Column | Meaning |
|---|---|
| Original target column | Original glucose values; originally missing values remain `NA`. |
| `TimeSeries` | Numeric time feature derived from the timestamp column. |
| `imputed_glucose_value` | Completed glucose values after imputation. |
| `imputation_method` | Final method used, such as `MICE+ARIMA` or `MICE+XGBoost`. |
| `missing_rate` | Original missingness rate of the target glucose column. |

# Previewing results

After imputation, the app displays a preview of rows where the original target
glucose value was missing. This is more informative than showing only the first
few rows of the completed data set because it lets users directly inspect the
newly imputed values.

For example, the preview is based on logic like:

```{r preview-logic, eval = FALSE}
imputed_rows <- out[is.na(out[[target_col]]), , drop = FALSE]
head(imputed_rows, 15)
```

The full completed data frame remains available for download.

# Downloading results

Use the **Download imputed CSV** button to save the completed data set. The CSV
contains all returned columns from `run_missing_glucose_imputation()`, including:

- the original glucose column;
- `TimeSeries`;
- `imputed_glucose_value`;
- `imputation_method`;
- `missing_rate`;
- any original input columns retained by the workflow.

Internal lag and rolling-mean columns are used during imputation but are removed
from the returned data frame before display or download.

# Troubleshooting

## The app does not launch

If you see an error saying that Shiny is not installed, run:

```r
install.packages("shiny")
```

Then restart R and try:

```r
run_cgmissingdata_app()
```

## No column choices appear

Column choices appear only after data are loaded. Upload a CSV file or load one
of the built-in example data sets.

## Imputation fails because a timestamp cannot be parsed

Check the timestamp column selected in the app. The values should be parseable
dates or datetimes, for example:

```r
"2020:01:16:00:00"
"2020-01-16 00:00:00"
"2020/01/16 00:00:00"
"2020-01-16T00:00:00"
```

If the wrong column was selected as the timestamp column, select the correct
column and rerun imputation.

## Python backend fails because a Python module is missing

If `imputer_backend = "sklearn"` fails because Python packages are missing, run:

```r
reticulate::py_require(c(
  "numpy",
  "pandas",
  "scikit-learn",
  "statsmodels",
  "xgboost"
))
```

Then restart R and launch the app again.

## Downloaded data contain `NA` in the original glucose column

This is expected. The original target column is intentionally preserved. The
completed values are stored in:

```r
imputed_glucose_value
```

# Developer notes

The recommended package structure for the app is:

```text
inst/
└── shiny/
    └── cgm_imputation_app/
        └── app.R
```

The launcher should live in an exported R function, for example:

```r
run_cgmissingdata_app <- function() {
  app_dir <- system.file(
    "shiny",
    "cgm_imputation_app",
    package = "CGMissingDataR"
  )
  shiny::runApp(app_dir, display.mode = "normal")
}
```

Because the app is optional, `shiny` should usually be listed in `Suggests`, not
`Imports`, unless the package requires Shiny for normal operation.
